### 2017-04-24 paper/jfp-submission/safe.tex
discuss speculative insantiation

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We found that for 247 programs (roughly a third of the failures, or 5\%
of all programs), \toolname got stuck when it tried to compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.
%
Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.

@@

To summarize, our investigation suggests that the majority of programs
(88\%) for which we fail to find a witness do not, in fact, admit a
witness.
%
Overall, it appears that our random instantiation of holes is well-suited
to finding type errors in student programs.
%
The remaining programs were all cases where \ocaml's type system was
overly conservative.
%
Of course, the conservatism is somewhat justified as each case pointed
to a ``code smell'' or difficult-to-use API; it would be interesting to
investigate how demonstrate these issues in a similar manner.

>>>>

\ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}

To answer this question, we investigated the 720 \ucsdbench programs for
which we failed to find a witness.
%
We found that for 247 programs (roughly a third of the failures, or 5\%
of all programs), \toolname got stuck when it tried to compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While extremely convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@ as input, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound, we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in the above @bad@ function, from type errors that were the fault of our
instantiation.

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, we can still view @x@ and @y@
\emph{morally} as holes, and thus treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.

Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

@@

To summarize, our investigation suggests that the majority of programs
(88\%) for which we fail to find a witness do not, in fact, admit a
witness.
%
Overall, it appears that our random instantiation of holes is well-suited
to finding type errors in student programs.
%
The remaining programs were all cases where \ocaml's type system was
overly conservative.
%
Of course, the conservatism is somewhat justified as each case pointed
to a ``code smell'' or difficult-to-use API;\ it would be interesting to
investigate how demonstrate these issues in a similar manner.
